End-User Software Customization by Direct Manipulation of Tabular Data

### 2020-05-23 paper.tex
maybe i'm done

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we introduce \emph{table-driven customization}, a new
  way for end users to extend software by direct manipulation without
  doing traditional programming. We augment a user interface with a
  table view showing the structured data inside the application. When
  users edit the table, their changes are reflected in the original UI.
  This simple model accommodates a spreadsheet formula language and
  custom data editing widgets, providing enough power to implement a
  variety of useful extensions.

@@

We subscribe to MacLean et al's vision of a ``gentle slope''
\citep{maclean1990} free of such abrupt ``cliffs,'' where users should
only need to make minimal and incremental investments in skill to
achieve their desired customizations. We seek to contribute to this
gentle slope with a new method for customizing software via direct
manipulation, which requires less skill than scripting but still
supports powerful customizations. We take inspiration from visual
database query interfaces and spreadsheets, which have successfully
enabled millions of end users to compute with data through direct
manipulation.

In our proposed paradigm, \emph{table-driven customization}, an
application's UI is augmented with a table view where the user can see
and manipulate the application's internal data. These changes don't just
apply to the table; they also result in immediate changes to the
application's original user interface. The user can sort/filter data in
the UI, inject annotations, pull in related information from other web
services, and more, all using the table as a mediating interface.
Interacting with the table view feels like interacting with a familiar
spreadsheet, but results in customizing an existing application.

@@

Finanlly, in Section~\ref{sec:conclusion}, we conclude and describe some
opportunities for future work.

@@

\textbf{Adding estimated read times}: Next, the user decides to attempt
a more substantial customization: to add estimated read times to each
article, in order to prioritize reading deeper content.

@@

They can then filter the table to only contain rows where this formula
column contains \texttt{false}; links that the user has already visited
are hidden both from the table view and the original page. This is an
example of a customization that the original website could not have
implemented, since websites don't have access to the browser history for
privacy reasons. But by using Wildcard, the user could implement the
customization locally, without needing to expose their browser history
to Hacker News.

@@

The two tables are then combined into a single table for the end user to
view and edit. The \textbf{query engine} is responsible for creating
this combined view, and routing the user's edits back to the individual
table adapters. In this example, the query engine has joined the two
tables together by a shared ID column, and sorted the result by the
points column.

@@

The default option is to return a DOM element, in which case the generic
adapter extracts the text contents of the DOM element and casts them to
the type of the column. For example, if the scraping function returns a
\texttt{\textless{}div\textgreater{}} containing the string ``1.23'',
and the column has a numeric type, the generic adapter extracts the
number 1.23. The advantage of this approach is that the field can be
\emph{writable}---when the user edits the value in the table, the
generic adapter can simply overwrite the inner contents of the DOM
element to reflect the change. In practice, this is most useful for
editing the value of form input fields, although it's technically
possible to enable the editing of other fields (e.g., editing the title
of a post on Hacker News).

@@

With the advent of rich frontend web frameworks, structured application
state is now often available in the web client. We suspect it is
possible to create plugins for frontend frameworks that expose this
state to Wildcard with only minimal effort from the application
developers. To test this hypothesis, we have created an early
proof-of-concept of an integrated website adapter for the popular
\href{https://redux.js.org/}{Redux} Javascript state management library.
To set up this integrated adapter, the programmer would only need to
make a few minor changes to their application code:

\begin{itemize}
\tightlist
\item
  \emph{Exposing a table}: In Redux, app developers already represent
  the state of a user interface as a single object. To expose a table of
  data to Wildcard, the developer simply writes a function that maps the
  state object to a table.
\item
  \emph{Editing data values}: The developer writes handlers which define
  how edits to table records should map to UI actions. For example,
  editing the value of a todo might trigger an \texttt{EDIT\_TODO}
  event.
\item
  \emph{Sorting/filtering the UI}: The developer inserts an additional
  data processing step before the state object is rendered in the UI, so
  that Wildcard can apply sorting and filtering.
\item
  \emph{Annotations}: The developer adds additional React components
  managed by the Wildcard adapter into their UI tree, which render
  additional data as annotations.
\end{itemize}

@@

The query engine is also responsible for executing formulas. We have
built a small formula language resembling a spreadsheet formula
language, except that formulas automatically apply across an entire
column of data, and reference other column names instead of values in
specific rows. This is more convenient than needing to copy-paste a
formula across an entire column as in spreadsheets, and has worked for
all of the customizations we have built. Visual database query systems
like SIEUFERD \citep{bakke2016} use a similar approach to using formulas
in a structured table.

@@

We provide a basic table editor as the user interface on top of the
query engine. Our table editor is built with the Handsontable Javascript
library, which provides UI elements for viewing, editing, sorting, and
filtering a table.

@@

We developed site-specific adapters for 11 websites that we personally
use frequently, and then built customizations in Wildcard for those
websites. Table 1 summarizes these results, showing the number of lines
of code in the adapter for each site, and some example customizations we
created. Here we offer our reflections from these experiences using the
system, focused on two key questions:

@@

There are many customizations that are not possible to implement with
table-driven customization. Some of the limitations are specific to the
current implementation of the Wildcard extension, but many are more
fundamental to the general paradigm.

@@

Some of the challenges of writing a DOM scraping adapter are the same
ones as with writing normal web scraping code, but the more interactive
nature of Wildcard introduces additional challenges. One challenge is
registering appropriate event handlers to update the table data in
response to UI changes that happen after initial page load. Another
challenge is persisting updates to the DOM---some websites use virtual
DOM frameworks that can occasionally overwrite changes made by Wildcard.
So far, in practice we've managed to work around these issues for all of
the websites we've tried, but we don't claim that any website can be
customized through DOM scraping. As web frontend code gets increasingly
complex (and starts to move beyond the DOM to other technologies like
Shadow DOM or even WebGL), it may become increasingly difficult to
customize websites from the outside without first-party support.

AJAX scraping proved highly useful in several cases. The Uber Eats
website was challenging to scrape because it has a complex DOM structure
with machine-generated CSS classes, but the site also uses AJAX requests
which contain all the relevant data in a structured form that is much
easier to extract. We also found examples where relevant information
wasn't present in the DOM at all. On the grocery delivery site
Instacart, we found that AJAX requests contained additional data not
shown in the UI, enabling us to do things like sort grocery items by
category.

@@

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotations by clicking and typing on the relevant row. At every
step, the user receives intermediate feedback, not only in the table
view, but also in the original application, so it's clear whether they
are making progress towards their desired result. These types of
interactions are common in GUI applications, and Wildcard therefore
seems to meet MacLean et al's goal: some one-click customizations are as
easy as using the original application. Formulas introduce some
additional complexity, but spreadsheets have demonstrated that formula
programming is still accessible to many users, due to the pure
functional semantics and the intermediate visibility of results.

One aspect of directness that we have chosen not to pursue in Wildcard
is enabling customization in closer proximity to the original user
interface elements, as explored by tools like Scotty \citep{eagan2011}.
We have found that augmenting the original UI with a distinct,
additional representation provides a more consistent experience across
all applications, and clearly shows the user what structured data is
available to work with.

@@

Our work is also inspired by the many projects that have explored using
spreadsheets as a foundation for building software applications,
including Object Spreadsheets \citep{mccutchen2016}, Quilt
\citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
\href{https://www.glideapps.com/}{Glide}. We share the main idea of
connecting a spreadsheet view to a GUI, but we apply it to software
customization, rather than building software from scratch.

@@

So far, most usage of Wildcard has come from members of the project
team. Our primary goal for future work is to evaluate the system with a
broader group of users. What usability barriers do end users face when
using the system? What types of customizations do they choose to create?
What formulas prove most helpful? Through real deployment of the tool,
we hope to validate the viability of creating a community-maintained
library of site-specific adapters.

@@

As computing plays an ever greater role in our lives, it is increasingly
important that end users have agency over the behavior of their
software, rather than having every detail be dictated by companies whose
incentives are not always aligned with the user's interests. We hope
that table-driven customization can serve as one point, and maybe a
guidepost, on our path from normal use to deep modification, in support
of a more adaptable experience for all computer users.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we introduce \emph{table-driven customization}, a new
  way for end users to extend software by direct manipulation without
  doing traditional programming. We augment existing user interfaces
  with a table view showing the structured data inside the application.
  When users edit the table, their changes are reflected in the original
  UI. This simple model accommodates a spreadsheet formula language and
  custom data editing widgets, providing enough power to implement a
  variety of useful extensions.

@@

We subscribe to MacLean et al's vision of a ``gentle slope''
\citep{maclean1990} free of such ``cliffs,'' where users should only
need to make minimal and incremental investments in skill to achieve
their desired customizations. We seek to contribute to this gentle slope
with a new method for customizing software via direct manipulation,
which requires less skill than scripting but still supports powerful
customizations. We take inspiration from visual database query
interfaces and spreadsheets, which have successfully enabled millions of
end users to compute with data through direct manipulation.

In our proposed paradigm, \emph{table-driven customization}, an
application's UI is augmented with a table view where the user can see
and manipulate the application's internal data. These changes don't just
apply to the table; they also result in immediate changes to the
application's original user interface. The user can sort/filter data in
the UI, inject annotations, pull in related information from other web
services, and more, all using the table as a mediating interface.
Interacting with the table view resembles interacting with a familiar
spreadsheet, but results in customizing an existing application.

@@

Finally, in Section~\ref{sec:conclusion}, we conclude and describe
opportunities for future work.

@@

\textbf{Adding estimated read times}: Next, the user decides to attempt
a more substantial customization: adding estimated read times to each
article, in order to prioritize reading deeper content.

@@

They can then filter the table to only contain rows where this formula
column contains \texttt{false}; links that the user has already visited
are hidden both from the table view and the original page. This is an
example of a customization that the original website could not have
implemented, since websites don't have access to the browser history for
privacy reasons. But by using Wildcard, the user was able to implement
the customization locally, without needing to expose their browser
history to Hacker News.

@@

The two tables are then combined into a single table for the user to
view and edit. The \textbf{query engine} is responsible for creating
this combined view, and routing the user's edits back to the individual
table adapters. In this example, the query engine has joined the two
tables together by a shared ID column, and sorted the result by the
points column.

@@

The default option is to return a DOM element, in which case the generic
adapter extracts the text contents of the DOM element and casts them to
the type of the column. For example, if the scraping function returns a
\texttt{\textless{}div\textgreater{}} containing the string ``1.23'',
and the column has a numeric type, the generic adapter extracts the
number 1.23. The advantage of returning a DOM element is that the value
is editable---when the user changes the value in the table, the generic
adapter can simply overwrite the inner contents of the DOM element. In
practice, this is most useful for editing the value of form input
fields. It's technically possible to edit the contents of any element in
the page (e.g.~editing the title of a post on Hacker News), but this
often isn't particularly useful, so adapter authors have the ability to
mark fields as readonly.

@@

With the advent of rich frontend web frameworks, structured application
state is now often available in the web client. We suspect it is
possible to create plugins for frontend frameworks that expose this
state to Wildcard with only minimal effort from the application
developers. To test this hypothesis, we created an early
proof-of-concept of an integrated website adapter for the popular
\href{https://redux.js.org/}{Redux} Javascript state management library.
In Redux, app developers conveniently already represent the state of a
user interface as a single object, so to expose a table of data to
Wildcard, the developer simply writes a function that maps the
centralized state object to a table.

@@

The query engine is also responsible for executing formulas. We have
built a small formula language resembling a spreadsheet formula
language. As in visual database query tools like SIEUFERD
\citep{bakke2016} and \href{https://airtable.com/}{Airtable}, formulas
automatically apply across an entire column of data, and reference other
column names instead of values in specific rows. This is more convenient
than needing to copy-paste a formula across an entire column as in
spreadsheets, and has worked for all of the customizations we have
built.

@@

We provide a table editor view as the user interface on top of the query
engine. Our table editor is built with the Handsontable Javascript
library, which provides built-in UI elements for viewing, editing,
sorting, and filtering a table.

@@

We developed site-specific adapters for 11 websites that we personally
use frequently, and then built customizations for those websites using
the Wildcard table view. Table 1 summarizes these results, showing the
number of lines of code in the adapter for each site, and some example
customizations we created. Here we offer our reflections from these
experiences using the system, focused on two key questions:

@@

There are many customizations that are not possible to implement with
table-driven customization. Some of the limitations are specific to the
current implementation of the Wildcard extension, but others are more
fundamental to the general paradigm.

@@

Some of the challenges of writing a DOM scraping adapter are the same
ones as with writing normal web scraping code. Sometimes, addressing the
desired set of elements can be difficult, and when sites change,
scrapers can break---we observed several instances where sites changed
their CSS classes and caused Wildcard adapters to no longer work. One
benefit of a library of shared wrappers is that if many customizations
depend on some piece of scraping logic, rather than having the scraping
logic embedded in a single browser extension, it should be more likely
to be fixed quickly.

The interactive nature of Wildcard also introduces additional challenges
beyond normal web scraping. One challenge is registering appropriate
event handlers to update the table data in response to UI changes that
happen after initial page load. Another challenge is persisting updates
to the DOM---some websites use virtual DOM frameworks that can
occasionally overwrite changes made by Wildcard. So far, in practice
we've managed to work around these issues for all of the websites we've
tried, but we don't claim that any website can be customized through DOM
scraping. As web frontend code gets increasingly complex (and starts to
move beyond the DOM to other technologies like Shadow DOM or even
WebGL), it may become increasingly difficult to customize websites from
the outside without first-party support.

AJAX scraping proved very useful in several cases. The Uber Eats website
was challenging to scrape because it has a complex DOM structure with
machine-generated CSS classes, but the site also uses AJAX requests
which contain all the relevant data in a structured form that is much
easier to extract. We also found examples where relevant information
wasn't present in the DOM at all. On the grocery delivery site
Instacart, we found that AJAX requests contained additional data not
shown in the UI, enabling us to do things like sort grocery items by
category.

@@

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotations by clicking and typing on the relevant row. At every
step, the user receives intermediate feedback, not only in the table
view, but also in the original application, so it's clear whether they
are making progress towards their desired result. These types of
interactions are common in GUI applications, and Wildcard therefore
seems to meet MacLean et al's goal: some one-click customizations are as
easy as using the original application. Formulas introduce some
additional complexity, but spreadsheets have demonstrated that formula
programming is still accessible to many users, helped by the pure
functional semantics and the intermediate visibility of results.

One aspect of directness that we have chosen not to pursue in Wildcard
is enabling customization in closer proximity to the original user
interface elements, as explored by tools like Scotty
\citep{eagan2011}---all customization occurs in a separate panel next to
the original UI. While closer proximity might be helpful, we have found
that augmenting the original UI with a distinct, additional
representation provides a more consistent experience across all
applications, and clearly shows what structured data is available to
work with.

@@

Our work is also inspired by the many projects that have explored using
spreadsheets as a foundation for building software applications,
including Object Spreadsheets \citep{mccutchen2016}, Quilt
\citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
and \href{https://www.glideapps.com/}{Glide}. We share the main idea of
connecting a spreadsheet view to a GUI, but we apply it to software
customization, rather than building software from scratch.

@@

So far, most usage of Wildcard has come from members of the project
team. Our primary goal for future work is to evaluate the system with a
broader group of users. What usability barriers do end users face when
using the system? What types of customizations do they choose to create?
What formulas prove most helpful? Through real deployment of the tool,
we hope to validate the viability of a community-maintained library of
site-specific adapters.

@@

As computing plays an ever greater role in our lives, it is increasingly
important that end users have agency over the behavior of their
software, rather than having every detail be dictated by companies whose
incentives are not always aligned with the user's interests. We hope
that table-driven customization can serve as a point on the path from
normal use to deep modification, in support of a more adaptable
experience for all computer users.
